Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2025-03-07 paper/markdown/rebuttal_letter.Rmd

---
title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Rebuttal Letter"
author: "Department of Radiology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Reviewer 1

## Point 1

### Issue

Add results in the abstract as well.

### Response

## Point 2

### Issue

The introduction is not sufficient.

### Response

## Point 3

### Issue

The literature review part needs to be added.

### Response

## Point 4

### Issue

After adding the literature review, add a summary table and clearly mention this study's research gaps and contributions.

### Response

## Point 5

### Issue

More details, such as features, are required in the study data.

### Response

## Point 6

### Issue

The paper lacks details about hyper-parameters for machine learning models. 
For example, what was the C and class weight value for the SVM model?

### Response

## Point 7

### Issue

Have you considered the GridSearchCV for model hyper-parameters or set it manually?

### Response

## Point 8

### Issue

Comparison with state-of-the-art studies is missing.

### Response

## Point 9

### Issue

The Practical Implication of the study is missing.

### Response

## Point 10

### Issue

Conclusion also not well presented.

### Response

# Reviewer 2

## Point 11

### Issue

The introduction does not sufficiently contextualize how other published multi-parameter or machine learning approaches have succeeded or failed. What specific gaps remain? A more detailed table is needed to perform the literature review.

### Response

## Point 12

### Issue

What is new or improved about your multi-parameter modeling approach compared to existing frameworks?

### Response

## Point 13

### Issue

More details about selection and potential biases (e.g., dropouts, missing visits) are needed for the data set.

### Response

## Point 14

### Issue

It remains unclear how each scoring system aligns, differs, or complements the other.

### Response

Overemphasis on “satisfactory accuracy” without deeper clinical context.

## Point 15

### Issue

The neural network model performed poorly for DLCO.

### Response

## Point 16

### Issue

The SHAP analysis identifies “top 15” predictors. However, the manuscript does not address correlations among those predictors or discuss whether certain features might be redundant.

### Response

## Point 17

### Issue

With “opacity cutoff: 0.12%” or “high opacity cutoff: 0.002%” appear extremely small. The clinical significance of such minimal changes in lung volume being “abnormal” is unclear—are these thresholds truly meaningful in practice?

### Response

## Point 18

### Issue

The paper concedes “model overfitting” could be an issue. However, it lacks any systematic attempt (e.g., a learning-curve analysis or thorough regularization strategy) to demonstrate that overfitting is minimal or under control.

### Response

# References

### 2025-03-07 paper/markdown/manuscript.Rmd
DLCO <80% predictions and DLCO values
Analysis of predictions of DLCO < 80% in observations with increasing DLCO values expressed as percentages of the patient's reference. Drop-out rates and frequency of LFT and CT abnormalities included in the manuscript text.

The longitudinal observation CovILD study cohort consisted of n = `r length(covild$common_ids)` convalescents of symptomatic SARS-CoV-2 infection initially enrolled during the fist outbreak of the pandemic (March - June 2020) in three clinical centers in western Austria [@Sahanic2023; @Luger2022; @Sonnweber2022; @Sonnweber2020]. 
Out of them, n = `r length(covild$complete_ids)` participants with n = `r nrow(covild$ct)` observations with complete CT and LFT data obtained at follow-up visits at the two, three, six and twelve months after COVID-19 diagnosis were included in the current analysis. 
Please note, that per participant, up to four observations were available. 
Due to this participant-matching, the observations were not independent, which had consequences for our analysis strategy (__Figure \@ref(fig:fig-consort)__). 
The majority of analyzed data points were obtained at the two- (n = `r covild$ct %>% filter(follow_up == '2-month follow-up') %>% nrow`) and three-month follow-up (n = `r covild$ct %>% filter(follow_up == '3-month follow-up') %>% nrow`, __Supplementary Table S1__ - __S2__).

@@

Frequency of any radiological lung abnormalities, GGO, reticulation, consolidation, and bronchiectasis diagnosed by a human radiologist was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients. 
GGO and reticulation were the most frequent lesion types, independently of the follow-up and acute disease severity. 
Resolution of radiological lung findings. 
Consolidation and bronchiectasis lesions were detected primarily in moderate and severe COVID-19 survivors at the two-month follow-up examination. 
The increased frequency and severity of structural lung lesions in hospitalized patients as well as gradual resolution of the findings was also evident from analysis of the time course of human-determined CTSS, and AI-measured lung opacity and high opacity (__Supplementary Figures S1__ and __S2__, __Supplementary Table S3__). 
As demonstrated by correspondence analysis, in the majority of observations with GGO, reticulations were also present (`r covild$ct %>% filter(GGO_finding == 'yes', reticulation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with GGO). 
consolidations co-occurred with GGO lesions only in a minute fractions of observations (`r covild$ct %>% filter(GGO_finding == 'yes', consolidation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with  GGO), most frequently at the two-month follow-up (__Supplementary Figure S3A__). 
A high degree of correlation was observed between human-determined CTSS, and AI-determined lung opacity and high opacity (Spearman's $\rho$: `r corr_lft$test$ct$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S3B__).

Insufficient DLCO, FVC, and FEV1 defined as values below < 80% of the patient's reference value were the most common abnormalities of lung function. 
As with the CT findings, their frequency was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients, independently of the follow-up time point. 
Additionally, lung function recovery reflected by decreasing rates of insufficient DLCO and FVC, and by an improvement of DLCO and FVC values expressed as percentages of the patient's reference was the most evident in severe COVID-19 survivors (__Supplementary Figures S4__ and __S5__, __Supplementary Table S4__). 
In most of observations with insufficient FVC, insufficient FEV1 was found (`r covild$lft %>% filter(FVC_reduced == 'yes', FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(FVC_reduced == 'yes') %>% nrow` observations with reduced FVC). 
By contrast, in approximately half of observations with insufficient DLCO reduced FVC or FEV1 was diagnosed (`r covild$lft %>% filter(DLCO_reduced == 'yes') %>% filter(FVC_reduced == 'yes' | FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(DLCO_reduced == 'yes') %>% nrow` observations with insufficient DLCO, __Supplementary Figure S6A__). 
Accordingly, numeric values of FVC and FEV1 were strongly correlated (Spearman's $\rho$ = `r corr_lft$test$lft %>% filter(variable1 == 'FVC_percent', variable2 == 'FEV1_percent') %>% .$rho %>% signif(2)`), while correlation of DLCO with FVC or FEV1 was only moderate (Spearman's $\rho$: `r corr_lft$test$lft %>% filter(variable1 == 'DLCO_percent' | variable2 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S6B__). 
This may suggest that reduced DLCO in COVID-19 survivors reflects a type of functional lung pathology that is distinct from abnormalities measured by FVC or FEV1.

Impaired physical performance defined by ECOG > 0 and exertional dyspnea defined by mMRC > 0 were the most frequent persistent COVID-19-related symptoms of potential relevance for pulmonary function, and their rates declined during convalescence in all acute COVID-19 severity subsets.  
Notably, we could not observe any substantial differences in frequency of physical performance impairment, dyspnea, and persistent cough between severe, moderate and ambulatory COVID-19 convalescents (__Supplementary Figure S7__, __Supplementary Table S5__).

@@

Prediction of lung function deficits following COVID-19 with demographic, clinical and biochemical explanatory factors and characteristic of acute SARS-CoV-2 infection suffers from low accuracy [@Sonnweber2022]. 
Additionally, relevance of residual structural lung lesions and severity of lung damage in COVD-19 convalescents for pulmonary function is not entirely clear [@Luger2022; @Sahanic2023]. 
Herein, we sought to model the major LFT abnormalities, i.e. insufficient DLCO, FVC, and FEV1, and numeric readouts, i.e DLCO, FVC, FEV1 in the CovILD cohort with a broad panel of `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` independent variables including baseline demographic and clinical characteristic (e.g. gender, age, smoking history, comorbidity), characteristic of acute COVID-19 (e.g. severity, medication, hospitalization) and convalescence (follow-up time point, persistent symptoms), as well as presence and severity CT abnormalities (e.g. GGO, CTSS, AI-determined opacity and high opacity, __Supplementary Table S2__). 
To this end, we employed four popular machine learning algorithms, 
Random Forest [@Breiman2001], 
gradient boosted machines (GBM) [@Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], and 
support vector machines (SVM) [@Weston1998]. 
Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for model optimization (tuning, __Supplementary Table S6__). 
Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).

Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference could be modeled with satisfactory accuracy in the cross-validation setting independently of the machine learning algorithm (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Low Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__). 
In turn, models of insufficient FVC and FEV1 exhibited poor predictive performance in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).

As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with sufficiently low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__). 
Analogically, no meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).

@@

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S12__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.

As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO. 
Other influential explanatory variables were 
well characterized risk factors of severe COVID-19 (age, male sex, body mass index, pre-existing malignancy and cardiovascular disease), 
readouts of acute COVID-19 severity (severity class, WHO ordinal severity scale, hospitalization length, ICU treatment, anti-coagulant treatment, and weight change), 
smoking intensity (pack-years), as well as 
rating of physical performance impairment (ECOG) and 
exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S13__, __Supplementary Table S9__).

>>>>

The longitudinal observation CovILD study cohort consisted of n = `r length(covild$common_ids)` convalescents of symptomatic SARS-CoV-2 infection initially enrolled during the fist outbreak of the pandemic (March - June 2020) in three clinical centers in western Austria [@Sahanic2023; @Luger2022; @Sonnweber2022; @Sonnweber2020]. 
Out of them, n = `r length(covild$complete_ids)` participants with n = `r nrow(covild$ct)` observations with complete CT and LFT data obtained at follow-up visits at the two, three, six and twelve months after COVID-19 diagnosis were included in the current analysis. 
Please note, that per participant, up to four observations were available. 
Due to this participant-matching, the observations were not independent, which had consequences for our analysis strategy (__Figure \@ref(fig:fig-consort)__). 
The majority of analyzed data points were obtained at the two- (n = `r covild$ct %>% filter(follow_up == '2-month follow-up') %>% nrow`) and three-month follow-up (n = `r covild$ct %>% filter(follow_up == '3-month follow-up') %>% nrow`, __Supplementary Table S1__ - __S2__). 
As compared with the initial set of patients = 145 with acute infection in the CovILD study by Sonnweber et al. [@Sonnweber2020], the dropout rates were `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['2-month follow-up']`%, `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['3-month follow-up']`%, `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['6-month follow-up']`% and  `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['12-month follow-up']`% at the two-. three-, six- and twelve-month follow-up.

@@

Radiological lung abnormalities were present in 
`r signif(table(lft_globals$analysis_tbl$CT_findings)['yes']/sum(table(lft_globals$analysis_tbl$CT_findings)), 2) * 100`% of observations (n = `r table(lft_globals$analysis_tbl$CT_findings)['yes']`); 
with GGO (`r signif(table(lft_globals$analysis_tbl$GGO_finding)['yes']/sum(table(lft_globals$analysis_tbl$GGO_finding)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$GGO_finding)['yes']`) 
and reticulation (`r signif(table(lft_globals$analysis_tbl$reticulation_finding)['yes']/sum(table(lft_globals$analysis_tbl$reticulation_finding)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$reticulation_finding)['yes']`) as the most common lesions. 
Frequency of any radiological lung abnormalities, GGO, reticulation, consolidation, and bronchiectasis diagnosed by a human radiologist was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients. 
GGO and reticulation were the most frequent lesion types, independently of the follow-up and acute disease severity. 
Consolidation and bronchiectasis lesions were detected primarily in moderate and severe COVID-19 survivors at the two-month follow-up examination. 
The increased frequency and severity of structural lung lesions in hospitalized patients as well as gradual resolution of the findings was also evident from analysis of the time course of human-determined CTSS, and AI-measured lung opacity and high opacity (__Supplementary Figures S1__ and __S2__, __Supplementary Table S3__). 
As demonstrated by correspondence analysis, in the majority of observations with GGO, reticulations were also present (`r covild$ct %>% filter(GGO_finding == 'yes', reticulation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with GGO). 
consolidations co-occurred with GGO lesions only in a minute fractions of observations (`r covild$ct %>% filter(GGO_finding == 'yes', consolidation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with  GGO), most frequently at the two-month follow-up (__Supplementary Figure S3A__). 
A high degree of correlation was observed between human-determined CTSS, and AI-determined lung opacity and high opacity (Spearman's $\rho$: `r corr_lft$test$ct$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S3B__).

Insufficient DLCO (`r signif(table(lft_globals$analysis_tbl$DLCO_reduced)['yes']/sum(table(lft_globals$analysis_tbl$DLCO_reduced)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$DLCO_reduced)['yes']`), 
FVC (`r signif(table(lft_globals$analysis_tbl$FVC_reduced)['yes']/sum(table(lft_globals$analysis_tbl$FVC_reduced)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$FVC_reduced)['yes']`), and 
FEV1 (`r signif(table(lft_globals$analysis_tbl$FEV1_reduced)['yes']/sum(table(lft_globals$analysis_tbl$FEV1_reduced)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$FEV1_reduced)['yes']`) defined as values below < 80% of the patient's reference value were the most common abnormalities of lung function. 
As with the CT findings, their frequency was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients, independently of the follow-up time point. 
Additionally, lung function recovery was the most pronounced in severe COVID-19 survivors as reflected by decreasing rates of insufficient DLCO and FVC, and by an improvement of DLCO and FVC values expressed as percentages of the patient's reference (__Supplementary Figures S4__ and __S5__, __Supplementary Table S4__). 
In most of observations with insufficient FVC, insufficient FEV1 was found (`r covild$lft %>% filter(FVC_reduced == 'yes', FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(FVC_reduced == 'yes') %>% nrow` observations with reduced FVC). 
By contrast, in approximately half of observations with insufficient DLCO reduced FVC or FEV1 was diagnosed (`r covild$lft %>% filter(DLCO_reduced == 'yes') %>% filter(FVC_reduced == 'yes' | FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(DLCO_reduced == 'yes') %>% nrow` observations with insufficient DLCO, __Supplementary Figure S6A__). 
Accordingly, numeric values of FVC and FEV1 were strongly correlated (Spearman's $\rho$ = `r corr_lft$test$lft %>% filter(variable1 == 'FVC_percent', variable2 == 'FEV1_percent') %>% .$rho %>% signif(2)`), while correlation of DLCO with FVC or FEV1 was only moderate (Spearman's $\rho$: `r corr_lft$test$lft %>% filter(variable1 == 'DLCO_percent' | variable2 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S6B__). 
This may suggest that reduced DLCO in COVID-19 survivors reflects a type of functional lung pathology that is distinct from abnormalities measured by FVC or FEV1.

Impaired physical performance defined by ECOG > 0 (`r signif(table(lft_globals$analysis_tbl$impaired_performance_symptom)['yes']/sum(table(lft_globals$analysis_tbl$impaired_performance_symptom)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$impaired_performance_symptom)['yes']`) and 
exertional dyspnea defined by mMRC > 0 (`r signif(table(lft_globals$analysis_tbl$dyspnea_symptom)['yes']/sum(table(lft_globals$analysis_tbl$dyspnea_symptom)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$dyspnea_symptom)['yes']`) were the most frequent persistent COVID-19-related symptoms of potential relevance for pulmonary function, and their rates declined during convalescence in all acute COVID-19 severity subsets.  
Notably, we could not observe any substantial differences in frequency of physical performance impairment, dyspnea, and persistent cough between severe, moderate and ambulatory COVID-19 convalescents (__Supplementary Figure S7__, __Supplementary Table S5__).

@@

Prediction of lung function deficits following COVID-19 with demographic, clinical and biochemical explanatory factors and characteristic of acute SARS-CoV-2 infection suffers from low accuracy [@Sonnweber2022]. 
Additionally, relevance of residual structural lung lesions and severity of lung damage in COVD-19 convalescents for pulmonary function is not entirely clear [@Luger2022; @Sahanic2023]. 
Herein, we sought to model the major LFT abnormalities, i.e. insufficient DLCO, FVC, and FEV1, and numeric readouts, i.e DLCO, FVC, FEV1 in the CovILD cohort with a broad panel of `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` independent variables including baseline demographic and clinical characteristic (e.g. gender, age, smoking history, comorbidity), characteristic of acute COVID-19 (e.g. severity, medication, hospitalization) and convalescence (follow-up time point, persistent symptoms), as well as presence and severity CT abnormalities (e.g. GGO, CTSS, AI-determined opacity and high opacity, __Supplementary Table S2__). 
To this end, we employed four popular machine learning algorithms, 
Random Forest [@Breiman2001], 
gradient boosted machines (GBM) [@Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], and 
support vector machines (SVM) with radial kernel [@Weston1998]. 
Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Table S6__). 
Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).

Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference yielded meaningful models as evaluated by cross-validation; accuracy of the models of DLCO < 80% was reproducible between the algorithms (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Low cross-validated Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__). 
In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).

As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__). 
No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).

@@

Prediction of reduced DLCO < 80% was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference by moving averages of accuracy and error metered as squared distance to the 0/1-coded outcome. 
As presented in __Supplementary Figure S12__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference. 
An analogical phenomenon was observed for the remaining machine learning algorithms (not shown). 
This illustrates that while predictions of highly compromised and normal-to-high DLCO are accurate, forecasts for borderline cases of reduced DLCO suffer from high error. 
This may question biological and clinical relevance of the 80% cutoff widely used for diagnosis of insufficient DLCO in COVID-19 survivors [@Sonnweber2020; @Zhang2022a].

To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO. With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes. 
The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
The learning curves for the best-performing GBM models are presented in __Supplementary Figure S13__, for the remaining algorithms comparable results were obtained (not shown). 
As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e. poor generalizability for unseen data. 
The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.

As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO. 
Other influential explanatory variables were 
well characterized risk factors of severe COVID-19 (age, male sex, body mass index, pre-existing malignancy and cardiovascular disease), 
readouts of acute COVID-19 severity (severity class, WHO ordinal severity scale, hospitalization length, ICU treatment, anti-coagulant treatment, and weight change), 
smoking intensity (pack-years), as well as 
rating of physical performance impairment (ECOG) and 
exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S14__, __Supplementary Table S9__).

Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S15A__). 
In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated. 
CTSS values were significantly higher in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`) (__Supplementary Figure S3B__, __Supplementary Table S10__). 
Analogically, significantly higher opacity was detected in observations with than without 
GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$eff_size`) and 
reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`). 
Effect sizes of these differences were large (__Supplementary Figure S15B__, __Supplementary Tables S11__).
Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
Tackling this redundancy e.g. by regularization may control the over-fitting and improve accuracy of predictions of DLCO and reduced DLCO by machine learning algorithms.
